---
title: "LabAge revisions"
output: html_notebook
---
Glen Oct 2025

```{r}
library(ggplot2)
library(gridExtra)
pregAgingDir = "C:/PregAging-master" #enter your PregAging-master directory location here
outputDir = sprintf("%s/revisions",pregAgingDir)
```

```{r}
data = read.csv(sprintf("%s/csvs/Lab distribution by age.csv",pregAgingDir))
```


#variable trajectories and population stats
you must run the jupyter notebook to produce sampled_data_allvars.csv
```{r}
ronData = read.csv(sprintf("%s/processed_data/sampled_data_allvars.csv",pregAgingDir),row.names=1) #random sample from healthy clalit
#ronData = read.csv(sprintf("%s/processed_data/sampled_data.csv",pregAgingDir),row.names=1) #random sample from healthy clalit #after feature selection
ronVars = sort(setdiff(colnames(ronData),"age"))

codex = read.csv(sprintf("%s/csvs/Metadata.csv",pregAgingDir))
codex=codex[!duplicated(codex[,"Test"]),]
rownames(codex)=codex[,"Test"]

codex[codex[,"units"]=="","units"] = "unknown"

exVars = c("PH_u","SPECIFIC_GRAVITY",
	          "HYPERperc","HYPO_perc",
	          "MACROperc",
	          "MICROperc_HYPOperc","MICR_perc",
	          "PCT"
           ) #excludes in original paper (only immune % are excluded in revised)
allVars= unique(c(ronVars,exVars))
```

# population stats
```{r}
print("age stats:")
quantile(ronData[,"age"])
mean(ronData[,"age"])

print("bmi stats:")
quantile(ronData[,"BMI"])
mean(ronData[,"BMI"])
```

# age dependencies
```{r}
g = list()
temp=ronData
for (j in 1:length(allVars))
{
  temp[,"y"] = ronData[,allVars[j]]
  g[[j]] = ggplot(temp,aes(x=age,y=y))+
    geom_point(size=.1)+
    geom_smooth()+
    labs(x="Age",y=sprintf("%s (%s)",codex[allVars[j],"Short.name"],codex[allVars[j],"units"]),title=codex[allVars[j],"Short.name"])+
    #scale_y_log10()+annotation_logticks(sides="l")+
    theme_minimal(base_size=12)#+
    #theme(title = element_text(size=5))
}

g[[1]]
#marrangeGrob(g,nrow=8,ncol=8,top=NULL)

save=T
if(save)
{
  layout=matrix(c(1:length(g),rep(NA,10^2-length(g))),nrow=10,ncol=10,byrow = T)
  ggsave(sprintf("%s/data_by_decade_allvars.png",outputDir),marrangeGrob(g,nrow=10,ncol=10,top=NULL,layout_matrix = layout),width=16,height=16)
}
```

#variable (feature) selection
-data from NHANES, ~1300 pregnancies plus age-matched controls using matchit package
  -https://wwwn.cdc.gov/nchs/nhanes/
-note: p values are just z-test of delta/deltase
```{r}
#library(scico) #nice colour palette
colors = c("1"="#7E1700", "2"="#C0E9C2", "3"= "#023198") #from scico roma palette

agg=read.csv(sprintf("%s/nhanes_labage_delta_preg_vs_control.csv",outputDir))
aggt=read.csv(sprintf("%s/nhanes_labage_delta_preg_vs_control_by_trimester.csv",outputDir))
aggt[,"trimester"] = ordered(aggt[,"trimester"])
#by system
agg[,"shortName"] = factor(agg[,"shortName"],unique(aggt[sort.list(-abs(aggt[,"delta"])),"shortName"]))
aggt[,"shortName"] = factor(aggt[,"shortName"],unique(aggt[sort.list(-abs(aggt[,"delta"])),"shortName"]))
g = list()
unsystems = unique(agg[,"system"])
ylim = range(c(abs(aggt[,"delta"])-aggt[,"deltase"],abs(aggt[,"delta"])+aggt[,"deltase"]),na.rm=T)
agg[,"p"] = 2*(1-pnorm((abs(agg[,"delta"]))/agg[,"deltase"]))
aggt[,"p"] = 2*(1-pnorm((abs(aggt[,"delta"]))/aggt[,"deltase"]))
for (i in 1:length(unsystems))
{
  g[[i]] = ggplot(subset(aggt,system==unsystems[i]),aes(x=shortName,y=abs(delta),ymin=abs(delta)-deltase,ymax=abs(delta)+deltase,shape=cut(p,c(-Inf,0.05,Inf),c("p < .05","NS"))))+
  geom_hline(yintercept=0,colour="grey",size=3)+
  geom_pointrange(data=subset(agg,system==unsystems[i]),colour="black")+ #overall
  geom_pointrange(aes(colour=trimester),position=position_dodge(.1))+ #by trimester
  scale_colour_manual(values=colors)+
  scale_y_continuous(limits=ylim)+
  labs(x="",y=bquote("|"*Delta*" | (years/sd)"),title=unsystems[i],shape="",colour="trimester")+
  #coord_flip()+
  theme_minimal(base_size=16)+
    theme(legend.position=c(.8,.7))
}
for (i in 2:length(g)) g[[i]] = g[[i]]+theme(legend.position = "none")
g[[1]]

save=T
if(save)
{
  ggsave(sprintf("%s/nhanes_variable_selection.png",outputDir),marrangeGrob(g,nrow=3,ncol=3,top=NULL),width=16,height=14)
}
```

#Least squares estimates
```{r}
#functions that will import and use donor correlations
LMFromCorEasy = function(C, #centered correlation matrix including x and y (not intercept)
                     N, #number of data points
                     mu, #mean of each variable (x and y) #NAMED
                     std, #standard deviation of each variable (x and y) #NAMED
                     ycol="y",
                     xcol=setdiff(colnames(C),ycol),
                     ridgePenalty=0,
                     ret="default"
                     )
{
  #basic easy version
  
  #warning("something is wrong here, Cv is too big")

  Cv = matrix(NA,nrow=nrow(C)+1,ncol=ncol(C)+1)
  Cv[1,1] = N/(N-1)
  Cv[1,-1] = mu
  Cv[-1,1] = mu
  Cv[1:nrow(C)+1,1:ncol(C)+1] = (C*outer(std,std)*(N-1)+N*outer(mu,mu))/(N-1)
  rownames(Cv) = c("(Intercept)",names(mu))
  colnames(Cv) = rownames(Cv)
  
  #print(range(Cv))
  #print(N)
  
  if(ret=="cov" | ret=="Cv") return(Cv)

  return(LMFromCov2(C=Cv,N=N,ycol=ycol,xcol=c("(Intercept)",xcol),unbiased=TRUE,ridgePenalty=ridgePenalty))
}

LMFromCov2 = function(C, #UNCENTERED covariance matrix #(unbiased estimator, so 1/(N-1) term)
                     N, #number of data points
                     ycol="y",
                     xcol=setdiff(colnames(C),ycol),
                     unbiased=TRUE, #did you use unbiased estimator for C?
                     ridgePenalty=0
                     )
{
  p = length(xcol)
  if(unbiased) C = C*(N-1)/N #de-unbias
  

  invCx = solve(C[xcol,xcol]+diag(ridgePenalty,length(xcol)))
  beta = invCx%*%C[xcol,ycol] #1/N*N cancel #this is right
  names(beta)=xcol
  
  #residual variance #don't really need N here but more accurate if we have
  s2 = N/(N-p)*(C[ycol,ycol]-c(C[ycol,xcol]%*%beta))
  
  #need N here
  beta_cov = s2*invCx/(N) #this is right
  dimnames(beta_cov)=list(xcol,xcol)
  
  return(list(beta=beta,se=sqrt(diag(beta_cov)),beta_cov=beta_cov,s=sqrt(s2),s2=s2))
}

SampleCor = function(vars,
                     systems,
                     rhoWithinSystemsDonors,
                     rhoBetweenSystemsDonors)
{
  #generate random covariance matrix based on within vs between system correlation structure
  C = matrix(NA,nrow=length(vars),ncol=length(vars))
  dimnames(C)=list(vars,vars)
  diag(C)=1
  unsystems=unique(systems)
  ind = 1
  for (j in 1:length(unsystems))
  {
    v = regVars[systems==unsystems[j]]
    v = intersect(v,colnames(C))
    notv = setdiff(colnames(C),v)
    if(length(v) < 1) next
    C[v,v]    = sample(rhoWithinSystemsDonors,size=length(v)^2,replace=T)
    C[v,notv] = sample(rhoBetweenSystemsDonors,size=length(v)*length(notv),replace=T)
    #C[notv,v] = C[v,notv]   #enforce symmetry #insufficient
  }
  #strict symmetry
  for (i in 1:length(vars))
  {
    for (j in i:length(vars))
    {
      C[j,i] = C[i,j]
    }
  }
  diag(C)=1
return(C)
}

SEM = function(x,na.rm=T)
{
  return(sd(x,na.rm=na.rm)/sqrt(sum(!is.na(x))))
}


```

you must run the jupyter notebook to save regression_data.csv
```{r}
regData = read.csv(sprintf("%s/processed_data/sampled_data_allvars.csv",pregAgingDir),row.names=1) 

regVars = ronVars

#nhanes feature selection:
regVars = c("PROTEIN_TOTAL_BLOOD", "CHOLESTEROL", "HEMOGLOBIN_A1C_CALCULATED","FIBRINOGEN",
         "PHOSPHATASE_ALKALINE", "UREA_BLOOD", "BPs","T3_FREE", "NEUT","TRANSFERRIN","RBC") #FERRITIN


regData = regData[,c("age",regVars)]
```

borrow NHANES correlation matrix
```{r}
Cref = read.csv(sprintf("%s/nhanes_female_covariance_reference_with_age_renamed2.csv",outputDir),row.names=1)
colnames(Cref)=rownames(Cref)
Cref= as.matrix(Cref)
Cref = cov2cor(Cref)
Nmat = read.csv(sprintf("%s/nhanes_female_covariance_reference_with_age_renamed2_N.csv",outputDir),row.names=1)
```

we will approximate missing correlation coefficients using within or between system mean (some things were never measured together e.g. T3_FREE and fibrinogen)
-this block is to get donors
```{r}
systems = codex[regVars,"Group"]
withinSystemDF = list()
betweenSystemDF =list()
unsystems=unique(systems)
ind = 1
for (j in 1:length(unsystems))
{
  v = regVars[systems==unsystems[j]]
  v = intersect(v,colnames(Cref))
  notv = setdiff(colnames(Cref),v)
  if(length(v) < 1) next
  Ctemp = Cref #correlation matrix
  Ctemp[lower.tri(Ctemp,diag=T)]= NA #remove redundant values
  withinSystemDF[[ind]] = data.frame(rho=c(Ctemp[v,v]),system=unsystems[j])
  betweenSystemDF[[ind]] = data.frame(rho=c(Ctemp[v,notv]),system=unsystems[j])
  ind = ind+1
}
withinSystemDF = do.call(rbind,withinSystemDF)
betweenSystemDF = do.call(rbind,betweenSystemDF)
withinSystemDF[,"type"]="within systems"
betweenSystemDF[,"type"]="between systems"
withinSystemDF = rbind(withinSystemDF,withinSystemDF)

rhoWithinDonors  = withinSystemDF[!is.na(withinSystemDF[,"rho"]),"rho"]
rhoBetweenDonors = betweenSystemDF[!is.na(betweenSystemDF[,"rho"]),"rho"] #very similar but no outliers
```


```{r}
v = c("age",regVars)
C0 = matrix(NA,nrow=length(v),ncol=length(v))
diag(C0)=1
colnames(C0)=v
rownames(C0)=v
v = intersect(rownames(Cref),rownames(C0))
C0[v,v] = Cref[v,v]
rhoDonors = Cref[v,v][upper.tri(Cref[v,v],diag=F)]
#rhoAgeDonors = Cageref[,1]
rhoAgeDonors=NA
C0["age",colnames(C0)] = cor(regData[,"age"],regData[,colnames(C0)],use='pairwise.complete')[1,]
C0[,"age"] = C0["age",]

if(any(is.na(C0[,"age"])) | any(is.na(C0["age",]))) stop("NAs in age dependence")
```

least-squares estimators
-goal is to see if base estimates are biased due to missing correlation structure (we sample from marginal distributions, thus ignoring correlations)
-we use donor correlation matrix from NHANES to approximate missing correlation structure and see how much it affects predictions (not much)
```{r}
#set.seed(1234)
mu = apply(regData,2,mean,na.rm=T)
names(mu)=colnames(regData)
std = apply(regData,2,sd,na.rm=T)
names(std)=colnames(regData)

#baseline estimate using least-squares
  #regData ignores correlations
m0 = lm(age~.,regData[,c("age",regVars)])
se = NA*coef(m0)
names(se)=names(coef(m0))
se[rownames(summary(m0)$coefficients)] = summary(m0)$coefficients[,"Std. Error"]
df0 = data.frame(var=names(coef(m0)),beta=coef(m0),se=se)[-1,]
rownames(df0)=df0[,"var"]
pr0 = predict(m0,regData,se.fit=TRUE)
pr020 = list()
pr020[[1]] = mean(pr0[[1]][regData[,"age"]==20],na.rm=T)
pr020[[2]] = SEM(pr0[[1]][regData[,"age"]==20],na.rm=T)


#estimate when correlation matrix is known
C = C0
Csample = SampleCor(vars=regVars,systems=codex[regVars,"Group"],rhoWithinSystemsDonors = mean(abs(rhoWithinDonors)), rhoBetweenSystemsDonors=mean(abs(rhoBetweenDonors)))
Csample[regVars,regVars] = Csample[regVars,regVars]*outer(sign(C0["age",regVars]),sign(C0["age",regVars])) #align sign to aging
C[regVars,regVars][is.na(C[regVars,regVars])] = Csample[regVars,regVars][is.na(C[regVars,regVars])]
N = mean(Nmat[regVars,regVars][upper.tri(Nmat[regVars,regVars])]) #rough estimate
m = LMFromCorEasy(C,mu=mu,std=std,N = N,ycol="age")
mcomb = m
Ccomb=C
Cvcomb = LMFromCorEasy(C,mu=mu,std=std,N = N,ycol="age",ret="cov")
Cvcomb = cbind(Cvcomb,N=NA)
df = data.frame(var=rownames(m$beta),beta=m$beta[,1],se=m$se)[-1,]
pr = as.matrix(regData[,regVars])%*%m$beta[regVars]+m$beta["(Intercept)"]
pr20 = mean(pr[regData[,"age"]==20],na.rm=T)


plotdf = rbind(data.frame(age=regData[,"age"],pr=pr,pr20=pr20,se=NA,model="cor"),
               data.frame(age=regData[,"age"],pr=pr0[[1]],pr20=pr020[[1]],se=pr0[[2]],model="no cor")
               )
g =ggplot(plotdf,aes(x=round(age),y=pr,colour=model))+
  #geom_abline(slope=1,intercept=0,colour="grey")+
  stat_summary(position = position_dodge(.1))+
  #labs(x="Age - 20",y=bquote(Delta~"= LabAge(age) - LabAge(20)"),colour="")+
  labs(x="Age ",y="LabAge",colour="")+
  theme_minimal(base_size=24)+
  theme(legend.position=c(.2,.8)
        #legend.background = element_rect(fill="white",colour=NA)
        )
g

save=T
if(save)
{
  #fit data
  lmdf = df[,"var",drop=F]
  lmdf[rownames(df0),"beta_nocor"] = df0[,"beta"]
  lmdf[rownames(df0),"beta_se_nocor"] = df0[,"se"]
  lmdf[rownames(df),"beta_cor"] = df[,"beta"]
  lmdf[rownames(df),"beta_se_cor"] = df[,"se"]
  
  write.csv(lmdf,sprintf("%s/nhanes_labage_fit.csv",outputDir),row.names=T)
  write.csv(mcomb$beta_cov,sprintf("%s/nhanes_labage_coefficient_variance.csv",outputDir),row.names=T)
  write.csv(Ccomb,sprintf("%s/nhanes_combined_correlations.csv",outputDir),row.names=T)
  write.csv(Cvcomb,sprintf("%s/nhanes_labage_sufficient_statistics.csv",outputDir),row.names=T)
  write.csv(Nmat[regVars,regVars],sprintf("%s/nhanes_labage_N.csv",outputDir),row.names=T)
  
  ggsave(sprintf("%s/labage_prediction_with_and_without_cor.png",outputDir),g,width=12,height=7,dpi=300)
}
```
